---
title: "msmtools"
author: "Francesco Grossetti"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{msmtools}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Package **msmtools** is an `R` package whose main goal is to facilitate the workflow with
longitudinal datasets which need to be analyzed in the context of multi-state models. In particular,
`msmtools` acts as the `msm` package companion.

## Longitudinal Dataset

Everytime we observe a given subject multiple times, we come up with a longitudinal dataset. This
means that measures are repeated $n$ times in a sequence which, in general, may not be equal
for all the subjects. Moreover, a longitudinal dataset could be viewed as a multilevel dataset:
a first level is given by the subject, and a second level is given by the single observation carried
out on that subject. A very common case of longitudinal dataset deals with hospital admissions.
A patient, our subject, can have a series of entries which correspond to hospital admissions. Each
hospital admission is recorded in a single row of the dataset. Let's consider a simplified version
ofthe `hosp` dataset which comes with `msmtools` package and represents synthetic
hospital admissions for 10 patients. For a detailed explanation of the dataset, please run `?hosp`.

```{r echo = FALSE}
library( msmtools )
data( hosp )
test = hosp[ 1:17, .( subj, adm_number, gender, age, label_2, dateIN, dateOUT, dateCENS ) ]
head( test, 17 )
```

As you can see, two patients are 'observed' 11 and 6 times through years, respectively.

These data format are very common when dealing with observational studies, or with chronic disease
monitoring. In general, they are a well stabilized system to collect information.


## Enhancing the Longitudinal Format with `augment()`

Why this structure is not enough if a multi-state model has to be run? A first observation could be
that we are not able to infer anything about the state in which a given subject (i.e. patient) is
at a particular point in time (i.e. hospital admission). The function `augment()` comes into play
for this reason: to take advantage of the longitudinal structure in order to extract usable
information to fuel a multi-state model. `augment()` takes a longitudinal dataset with exact
starting and ending time and reshape that to produce an 'augmented' version. For instance, applying
`augment()` to the dataset above produces the following dataset:


```{r echo = FALSE}
library( msmtools )
data( hosp )
hosp_augmented = augment( data = hosp, data_key = subj, n_events = adm_number, pattern = label_3,
t_start = dateIN, t_end = dateOUT, t_cens = dateCENS, verbose = F )
head( hosp_augmented[ , .( subj, adm_number, gender,
age, label_2, augmented, status, n_status ) ], 35 )
```












