---
title: "Building augmented data for multi-state models: the msmtools package"
subtitle: "*version 2.0.0*"
author: |
  | Francesco Grossetti
  | francesco.grossetti@unibocconi.it
date: "`r Sys.Date()`"
fontsize: 12pt
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
    toc_depth: 4
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Building augmented data for multi-state models: the msmtools package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r ninja, echo = FALSE }
library( msmtools )
library( ggplot2 )
```

***
# Introduction

Package **msmtools** is an `R` package which has been conceived to facilitate 
the workflow with longitudinal datasets which need to be analyzed in the context 
of multi-state models. In particular, **msmtools** acts as the **msm** package 
companion [@jackson2011multi; @msm_cran]. Moreover, **msmtools** is 
focused on efficiency and speed, also thanks to package `data.table` [@data.table].

## Longitudinal Dataset

Everytime we observe a given subject multiple times, we come up with a 
longitudinal dataset. This means that measures are repeated $n$ times in a 
sequence which, in general, may not be equal for all the subjects. Moreover, a 
longitudinal dataset could be viewed as a multilevel dataset: a first level is 
given by the subject, and a second level is given by the single observation carried
out on that subject. A very common case of longitudinal dataset deals with 
hospital admissions. A patient, our subject, can have a series of entries which 
correspond to hospital admissions. Each hospital admission is recorded in a 
single row of the dataset. Let's consider a simplified version of the `hosp` 
dataset which comes with **msmtools** package and represents synthetic
hospital admissions for 10 patients. For a detailed description of the dataset, 
please run `?hosp`. For demonstration purposes, we extract only the first 2 
patients, reducing the `hosp` dataset to a test sample of 17 rows per 8 
variables as you can see below.

```{r long_example, collapse = TRUE}
data( hosp )
hosp[ 1:17, .( subj, adm_number, gender, age, label_2, 
               dateIN, dateOUT, dateCENS ) ]
```

So, these two patients are 'observed' 11 and 6 times through years, respectively.

These data format are very common when dealing with observational studies, or 
with chronic disease monitoring and with hospital admissions recording.
In general, they are a well stabilized system to collect information.


## Enhancing the Longitudinal Structure with `augment()`

Why the standard longitudinal structure is not enough if a multi-state model 
has to be run? A first observation could be that we are not able to infer 
anything about the state in which a given subject (i.e. a patient) is at a 
particular point in time (i.e. hospital admission).
The function `augment()` comes into play for this reason: to take advantage of 
the longitudinal structure in order to extract usable information to fuel a 
multi-state model. `augment()` takes a longitudinal dataset with exact starting 
and ending times and reshape it to produce an *augmented* version. 
For instance, if you apply `augment()` to the dataset above, you get what 
follows:


```{r running_augment, collapse = TRUE}
hosp_augmented = augment( data = hosp, data_key = subj, 
                          n_events = adm_number, pattern = label_2,
                          t_start = dateIN, t_end = dateOUT, 
                          t_cens = dateCENS, verbose = FALSE )

hosp_augmented[ 1:35, .( subj, adm_number, gender, age, label_2, 
                         augmented, status, n_status ) ]
```

Despite the fact that not the same variables have been reported because of 
layout concerns, two things come up at first sight. In the first place, the 
number of rows is more than doubled. We now have 35 observations against the 
initial 17. In the second place, new variables have been created. We will 
describe them in a minute.

Given the complexity of the data, which can be very high, building a subject 
specific status flag which marks a its condition at given time steps, could be 
tricky and computationally intensive. At the end of the study, so at the 
censoring time, a subject, in general, can be alive, dead inside a given 
transition if death occurs within `t_start` and `t_end`, or outside a given 
transition if death occurs otherwise. After $n$ events, the corresponding flag 
sequence is given by $2n + 1$ for subjects alive and dead outside the transition, 
while it is just $2n$ for subjects who died inside of it. Let us consider an 
individual with 3 events. His/her status combinations will be as follows:

* **ALIVE**: IN-OUT | IN-OUT | IN-OUT | OUT 

* **DEAD OUT**: IN-OUT | IN-OUT | IN-OUT | DEAD

* **DEAD IN**: IN-OUT | IN-OUT | IN-DEAD.

This operation produces a dataset in the augmented long format which allows 
to neatly model transitions between the given states.

From now on, we refer to each row as a transition for which we define a state 
in which the subject lies. `augment()` automatically creates 4 new variables 
(if argument `more_status` is missing):

* *augmented*: the new timing variable for the process when looking at transitions. If
`t_augmented` is missing, then `augment()` creates *augmented* by default.
The function looks directly to `t_start` and `t_end` to build it and thus it 
inherits their class. In particular, if `t_start` is a date format, then 
`augment()` computes a new variable cast as integer and names it *augmented_int*. 
If `t_start` is a difftime format, then `augment()` computes a new variable 
cast as a numeric and names it *augmented_num*;

* *status*: a status flag which looks at `state`. `augment()` automatically 
checks whether argument `pattern` has 2 or 3 unique values and computes the 
correct structure of a given subject. The variable is cast as character;

* *status_num*: the corresponding integer version of *status*;

* *n_status*: a mix of *status* and `n_events` cast as character. *n_status* 
comes into play when a model on the progression of the process is intended.

### Working correctly with `augment()`

The main and only aim of `augment()` is data wrangling. When dealing with 
complex structures as longitudinal data are, it is really important to introduce 
some rules which help the user to not fail when using the function. Here we 
discuss some of these rules which are mandatory in order to get a correct 
workflow with `augment()`.

There are some arguments which are fundamental. They are `pattern` and `state`. 
`pattern` must contains the condition of a given subject at the end of the study. 
That is, how the subject is found at the censoring time. Because this peculiar 
structure is very common when dealing with hospital admission, the algorithm of 
`augment()` takes this framework as a reference. So, what does this mean? 
`pattern` can be either an integer, a factor or a character. Suppose we have it as 
an integer. `augment()` accepts only a pattern variable which have 2 or 3 unique 
values (i.e. running `length( unique( pattern ) )` must return 2 or 3). 
Now, suppose we provide a variable with 3 unique values. 
They must be 0, 1, and 3, nothing different than that. The coding for this is as
follows: 

1. Case 1. *integer*:

    + `pattern = 0`: subject is alive at the censoring time;
    + `pattern = 1`: subject is dead during a transition;
    + `pattern = 2`: subject is dead out of a transition.
    
1. Case 2. *factor*:

    + `pattern = 'alive'`: this is the first level of the factor and corresponds 
    to `pattern = 0` when integer;
    + `pattern = 'dead in'`: this is the second level of the factor and 
    corresponds to `pattern = 1` when integer;
    + `pattern = 'dead out'`: this is the third level of the factor and 
    corresponds to `pattern = 2` when integer;
    
1. Case 3. *character*:

    + the unique values must be in alphabetical order to resemble the pattern 
    of the integer and factor case.

In case one passes to `pattern` a variable which contains only two unique values, 
`augment()` automatically detects if the unit has an absorbing state occurred 
inside or outside a given transition. We suggest to always provide a variable 
with three unique values because of efficiency and speed.  
Everything else different from what described above will inevitably produce 
wrong behavior of `augment()` with and wrong results.

The second important argument is `state`. This is passed as a list and contains 
the status flags which will be used to compute all the status variables for the 
process. The length of `state` is 3, no less, no more and comes with a default 
given by: `state = list( 'IN', 'OUT', 'DEAD' )`. The order is important here too. 
The status flags must be passed such that the first one (`'IN'`)
represents the first state (in `hosp`, being inside a hospital), the second one 
represents the second state (in `hosp`, being outside a hospital), and the 
third one represents the absorbing state (in `hosp`, being dead inside or 
outside a hospital). As you can tell, this peculiar structure is typical of 
hospital admissions, where a patient can enter a hospital, can be discharged
from it, or can die. As we have already see, the `'DEAD'` status is reached no 
matter if the subject has died inside or outside a transition (i.e. in our case, 
inside or outside the hospital). One can change the flags in `state`, but they 
must be exactly 3. One may need a higher level of complexity when specifying 
the states of subjects. This will be discussed in the next section where `state` 
acts as the main indicator of states.

From version 1.2, `augment()` no longer checks the presence of missing data in 
the arguments passed to it. In fact, the new argument `check_NA` has been 
introduced and set to `FALSE` by default. Again, this is due because `augment()` 
has been developed just for restructuring data. The search for 
missing data is computationally intensive and could cause memory overheads. 
When dealing with very highly dimensional dataset, this becomes very unfeasible.
We then suggest to perform all these types of checks before running `augment()`. 
If one really wants to run this procedure, can set `check_NA = TRUE` and 
detection will be performed over `data_key`, `n_events`, `pattern`, `t_start` 
and `t_end`. Beware, that no missing imputation or deletion is carried out. 
If any missing value is found, then `augment()` throws an error asking 
you to fix the problem.

## What if a more complex status structure is needed?

`augment()` by default takes a very simple status structure given by 3 different 
values. In general, this is enough to define a multi-state model. But what if 
we needed a more complex structure. Let’s consider again the dataset `hosp` for 
the 3rd, 4th, 5th, and 6th patient with the following variables:

```{r subsetting, collapse = TRUE }
hosp[ 18:28, .( subj, adm_number, rehab, it, rehab_it,
                dateIN, dateOUT, dateCENS ) ]
```

As you can see, we have two variables which take into account the type of 
hospital admission. *rehab* marks a rehabilitation admission while *it* marks 
an intensive therapy one. They are both binary and integer variables, so one can 
compose them to get something which is informative and, at the same time, 
usable in the context of ‘making a status’. We then created the variable 
*rehab_it* which marks all the information in one place and it is a character.
You can pass *rehab_it* to the argument `more_status` to tell `augment()` to 
add these information into a new structure. Now, it is important to remember 
that `augment()` introduces some rules when you require to compute a more 
complex status structure. As you can see from the dataset, many values of 
*rehab_it* are set to `df`. This stands for ‘default’ and when `augment()` 
finds it, it just compute the default status you already passed to argument 
state (i.e. in this case, it can be ‘IN’, ‘OUT’, or ‘DEAD’). The argument 
`more_status` always looks for the value `df`, hence whenever you need to
specify a default transition make sure to label it with this value. 
So, if we run `augment()` on this sample, we obtain the following:

```{r complex_status, collapse = TRUE }
hosp_augmented_more = augment( data = hosp, data_key = subj,
                               n_events = adm_number, pattern = label_2,
                               t_start = dateIN, t_end = dateOUT,
                               t_cens = dateCENS, more_status = rehab_it,
                               verbose = FALSE )

hosp_augmented_more[ 36:60, .( subj, adm_number, rehab_it,
                               augmented, status, status_exp, n_status_exp ) ]
```

Beside the usual status variables, of which we reported only status, `augment()` 
computed two more:

* *status_exp*: is the direct expansion of status and the variable you passed 
to `more_status`, which in this case is *rehab_it*. The function composes them 
by pasting a `_` in between. This is the main reason why it is worth to build a 
character variable if you know you need to fuel it in as an indicator of a more 
complex status structure;

* *status_exp_num*: the corresponding integer version of *status_exp*;
* *n_status_exp*: similar to what has been done before, `augment()` mixes 
information coming from the current expanded status and the number of admission 
to give you the time evolution of the process.

## Polishing the augmented data

From version 1.3, **msmtools** gained the new function `polish()` which adds
support in the preprocessing phase. The function detects whether there are 
transitions which occur at the same time but land on different states within a
subject as specified by `data_key`. `polish()` needs to be run on an augmented 
dataset as computed by `augment()`. It then checks the presence of duplicates
trough the `time` parameter which refers to the new time variable for the process.
By default, `polish()` searches for either `augmented_int` or `augmented_num` 
depending on what `augment()` detected during the reshaping procedure.

***
# Graphical Assessment of a Multi-state Model

**msmtools** has been mainly developed to easily manage and work with 
longitudinal datasets that need to be restructured in order to get `msm()` to work 
properly. This is the main goal of the function `augment()`.

However, **msmtools** comes with two additional functions which try to graphically evaluate
the Goodness-of-Fit (GoF) of a multi-state model estimated via `msm()`. 
In this framework, GoF is always a tough quest. Furthermore, up to now, no formal statistical tests are defined 
when a multi-state model is computed within an exact time framework
[@titman2009model; @titman2008general].

## Comparing fitted and empirical survival with `survplot()`

One of the most common graphical method to assess whether a multi-state model 
is behaving properly, compares the empirical survival probability, as given by the Kaplan-Meier estimator,
with the estimated one. The function `survplot()` helps in this regards and does few more things.
The function is primarily a wrapper of the already known `plot.survfit.msm()` from the package **msm**.

Suppose we run a multi-state model on the included dataset `hosp` with the following code:

```{r multistate_model, collapse = TRUE }
# let's define the initial transition matrix for our model
Qmat = matrix( data = 0, nrow = 3, ncol = 3, byrow = TRUE )
Qmat[ 1, 1:3 ] = 1
Qmat[ 2, 1:3 ] = 1
colnames( Qmat ) = c( 'IN', 'OUT', 'DEAD' )
rownames( Qmat ) = c( 'IN', 'OUT', 'DEAD' )
Qmat

# attaching the msm package and running the model using
# gender and age as covariates
library( msm )
msm_model = msm( status_num ~ augmented_int,
                 subject = subj, data = hosp_augmented,
                 covariates = ~ gender + age,
                 exacttimes = TRUE, gen.inits = TRUE,
                 qmatrix = Qmat, method = 'BFGS',
                 control = list( fnscale = 6e+05, trace = 0,
                                 REPORT = 1, maxit = 10000 ) )
```

To run a comparison between the estimated and the empirical survival curves, we run the following
simple code:

```{r survplot_1, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
gof1 = survplot( x = msm_model, km = TRUE )
```

No surprise here, the plot is not very satisfying due to the really small dataset we used for
estimation. The workflow of `survplot()` has been greatly simplified and from version 2.0.0 it takes
fewer parameters. It also supports **ggplot2**. By default, `survplot()` returns a `gg/ggplot` 
object and always renders the plot requested. This allows the user to further manipulate the 
plot itself. I recommend to always assign the output to something, in this case `gof1`.  

`survplot()` can plot the survival probability for any valid transition as defined in the transition
matrix passed to `msm()`. For instance, the previous plot shows the survival curves for the transition 
(IN - DEAD). We can use the argument `from` and `to` to specify other transitions. 
If `to` is missing, `survplot()` checks what is the higher value in the
corresponding `msm` object and grabs it. Let's now plot the survival comparison for the transition 
(OUT - DEAD):

```{r survplot_2, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
gof2 = survplot( x = msm_model, km = TRUE, from = 2 )
```

If we do not want to show the Kaplan-Meier, just set `km = FALSE` to go back to the default behavior.

### Defining a custom time sequence

By default `survplot()` computes the estimated survival on a given grid. 
The number of grid points is given by `grid`. In some cases, one would like to 
explicitly pass a custom time sequence. This can be achieved by passing to the argument 
`times` a numeric vector. Now `grid` is ignored.

Consider our dataset and suppose we want to compute an estimated survival only for 
specific points in time. The following code addresses this request.

```{r custom_time_seq, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
time_seq = seq( 300, 800, by = 30 )
custom_time = survplot( x = msm_model, times = time_seq )
```

### Obtaining the dataset for the Kaplan-Meier

In addition to plotting, `survplot()` can return the desired data structure associated with a 
specific survival curve. For instance, by specifying `out = "km"` the data used to draw the 
Kaplan-Meier curve is returned. The code below shows how this works.

<!-- It is possible to tell `survplot()` to return the associated Kaplan-Meier  -->
<!-- dataset by setting `return.km = TRUE`. This fastly computes the data through  -->
<!-- `data.table`. Passing only `km = TRUE` won't return any data, even if they must  -->
<!-- be computed anyway to plot results. -->

```{r returnKM, collapse = TRUE, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
km_out = survplot( x = msm_model, km = TRUE, out = "km")
```

There are a couple of things to note. First, if one wants to get the data associated with the 
Kaplan-Meier, `km` must be set to `TRUE`. There is no possibility to get the data without plotting
the related empirical curve. Second, if `out` is not set to `"none"`, `survplot()` returns a named list.
In the first position there is always the plot identified by the element `$p`. In this case, the 
second element is the Kaplan-Meier idenfied by `$km`. You can see the `head()` below.

```{r}
head(km_out$km)
```

As you can see, the structure of the data is consistent. `survplot()` always returns a dataset 
with the following columns ordered by either `time` or `time_exact`.

* *subject*: the subject ID as passed to `msm()` function;

* *time*: the time at which the event occurred;

* *time_exact*: the relative time at which the event occurred. This is defined when `exacttimes = TRUE`;

* *anystate*: transition indicator to compute the Kaplan-Meier;

* *km*: the value of the empirical survival probability.


### Obtaining the dataset for the estimated survival

Similarly to what has been done for the Kaplan-Meier, it is possible to obtain the data 
used to compute the estimated survival as well. This can be achieve by setting 
`out = "fitted"`. When `times` is passed, the resulting dataset will have 
as many rows as the elements in `times`. If `times` is missing, then `survplot()` 
uses `grid` to know how many time points are requested. An example is shown in the next snippet.

```{r returnP, collapse = TRUE, fig.align='center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
fitted = survplot( x = msm_model, grid = 10, out = "fitted" )
```

As before, we have a named list with two elements, the first one is the plot and the second one
is the desired data structure which can be accessed with `$fitted`. You can see the `head()` below.

```{r}
head(fitted$fitted)
```

The structure of the data is consistent here too. `survplot()` always computes 
a dataset with two columns and it is ordered by time.

* *time*: time at which to compute the estimated survival. It can be obtained 
either by `grid` or by `times` so that the cardinality of the data depends on them;

* *surv*: the value of the estimated survival probability.

### Obtaining all the data

Of course, you can request `survplot()` to return both the datasets by specifying `out = "all"`.
In this case, a named list with three elements is returned. The mechanism to access the elements is
the same described above. The snippet below reports an example.

```{r return_all, collapse = TRUE, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
# just running survplot()
out_all = survplot( x = msm_model, km = TRUE, out = "all")
```

## Comparing expected and observed prevalences with `prevplot()`

A second graphical tool which helps us in the attempt to understand the 
goodness of the model is given by comparing the expected and observed prevalences [@gentleman1994multi]. `prevplot()` is a wrapper of the `plot.prevalence.msm()` 
function inside the **msm** package but, again, it does more things.

Consider the multi-state model we have built above. We can compute the 
prevalences using `prevalence.msm()` function. This produces a named list which 
will be used inside `prevplot()`. For instance, running the following code 
builds a plot of prevalences for each state of the model.

```{r prev, collapse = TRUE, fig.align = 'center', fig.width=7, fig.height=6, fig.fullwidth = TRUE}
# defining the times at which compute the prevalences
t_min = min( hosp_augmented$augmented_int )
t_max = max( hosp_augmented$augmented_int )
steps = 100L

# computing prevalences
prev = prevalence.msm( x = msm_model, covariates = 'mean', ci = 'normal',
                            times = seq( t_min, t_max, steps ) )

# and plotting them using prevplot()
prev_plot = prevplot( x = msm_model, prev.obj = prev, exacttimes = TRUE, M = FALSE, ci = TRUE )
```

It is mandatory for `prevplot()` to work with a `msm` object and a list 
computed by `prevalence.msm` are passed.

It is also possible to plot the following statistic:

$$
M = \frac{(O_{is} - E_{is})^2} {E_{is}}
$$
This gives an idea of the deviance from the Markov model and is computed 
according to Titman and Sharples (2008). The following code addresses this request.

```{r plot_M, collapse = TRUE, fig.align = 'center', fig.width=7, fig.height=9, fig.fullwidth = TRUE, warning = F}
gof = prevplot( x = msm_model, prev.obj = prev, exacttimes = TRUE, M = TRUE, ci = TRUE )
```



# References


