---
title: "Building augmented data for multi-state models: the `msmtools` package"
author: "Francesco Grossetti"
date: "`r Sys.Date()`"
fontsize: 12pt
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
    toc_depth: 4
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Using msmtools with longitudinal data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r ninja, echo = FALSE }
library( msmtools )
```

# Introduction

Package `msmtools` is an `R` package whose main goal is to facilitate the workflow with
longitudinal datasets which need to be analyzed in the context of multi-state models. In particular,
`msmtools` acts as the `msm` package companion [@jackson2011multi]. Moreover, `msmtools` is 
focused on efficiency and speed, also thanks to package `data.table` [@data.table].

## Longitudinal Dataset

Everytime we observe a given subject multiple times, we come up with a longitudinal dataset. This
means that measures are repeated $n$ times in a sequence which, in general, may not be equal
for all the subjects. Moreover, a longitudinal dataset could be viewed as a multilevel dataset:
a first level is given by the subject, and a second level is given by the single observation carried
out on that subject. A very common case of longitudinal dataset deals with hospital admissions.
A patient, our subject, can have a series of entries which correspond to hospital admissions. Each
hospital admission is recorded in a single row of the dataset. Let's consider a simplified version
of the `hosp` dataset which comes with `msmtools` package and represents synthetic
hospital admissions for 10 patients. For a detailed description of the dataset, please run `?hosp`.
For demonstration purposes, we extract only the first 2 patients, reducing the `hosp` dataset
to a test sample of 17 rows per 8 variables as you can see below.

```{r long example, collapse = TRUE}
data( hosp )
hosp[ 1:17, .( subj, adm_number, gender, age, label_2, 
               dateIN, dateOUT, dateCENS ) ]
```

So, these two patients are 'observed' 11 and 6 times through years, respectively.

These data format are very common when dealing with observational studies, or with chronic disease
monitoring and with hospital admissions recording.
In general, they are a well stabilized system to collect information.


## Enhancing the Longitudinal Structure with `augment()`

Why the standard longitudinal structure is not enough if a multi-state model has to be run?
A first observation could be that we are not able to infer anything about the state in which a given
subject (i.e. patient) is at a particular point in time (i.e. hospital admission).
The function `augment()` comes into play for this reason: to take advantage of the longitudinal
structure in order to extract usable information to fuel a multi-state model. `augment()` takes a
longitudinal dataset with exact starting and ending times and reshape it to produce an *augmented*
version. For instance, if you apply `augment()` to the dataset above, you get what follows:


```{r running augment, collapse = TRUE}
hosp_augmented = augment( data = hosp, data_key = subj, 
                          n_events = adm_number, pattern = label_3,
                          t_start = dateIN, t_end = dateOUT, 
                          t_cens = dateCENS, verbose = FALSE )

hosp_augmented[ 1:35, .( subj, adm_number, gender, age, label_2, 
                         augmented, status, n_status ) ]
```

Despite the fact that not the same variables have been reported because of layout concerns, two
things come up at first sight. In the first place, the number of rows is more than doubled. We now
have 35 observations against the initial 17. In the second place, new variables have been created.
We will describe them in a minute.

Given the complexity of the data, which can be very high, building a subject specific status flag
which marks a its condition at given time steps, could be tricky and computationally intensive.
At the end of the study, so at the censoring time, a subject, in general, can be alive, dead inside
a given transition if death occurs within `t_start` and `t_end`, or outside a given transition if
death occurs otherwise. After $n$ events, the corresponding flag sequence is given by $2n + 1$ for
subjects alive and dead outside the transition, while it is just $2n$ for subjects who died inside
of it. Let us consider an individual with 3 events. His/her status combinations will be as follows:

* **ALIVE**: IN-OUT | IN-OUT | IN-OUT | OUT 

* **DEAD OUT**: IN-OUT | IN-OUT | IN-OUT | DEAD

* **DEAD IN**: IN-OUT | IN-OUT | IN-DEAD.

This operation produces a dataset in the augmented long format which allows to neatly model
transitions between the given states.

From now on, we refer to each row as a transition for which we define a state in which the subject
lies. `augment()` automatically creates 4 new variables (if argument `more_status` is
missing):

* *augmented*: the new timing variable for the process when looking at transitions. If
`t_augmented` is missing, then `augment()` creates *augmented* by default. *augmented*.
The function looks directly to `t_start` and `t_end` to build it and thus it inherits their class.
In particular, if `t_start` is a date format, then `augment()` computes a new variable cast as 
integer and names it *augmented_int*. If `t_start` is a difftime format, then `augment()` computes
a new variable cast as a numeric and names it *augmented_num*;
* *status*: a status flag which looks at `state`. `augment()` automatically checks whether
argument `pattern` has 2 or 3 unique values and computes the correct structure of a given subject.
The variable is cast as character;
* *status_num*: the corresponding integer version of *status*;
* *n_status*: a mix of *status* and *status_num* cast as character. *status_num* comes into play
when a model on the progression of the process is intended.

### Working correctly with `augment()`

The main and only aim of `augment()` is data wrangling. When dealing with complex structures as
longitudinal data are, it is really important to introduce some rules which help the user to not fail
when using the function. Here we discuss some of these rules which are mandatory in order to get
a correct workflow with `augment()`.

There are some arguments which are fundamental. They are `pattern` and `state`. `pattern` must 
contains the condition of a given subject at the end of the study. That is, how the
subject is found at the censoring time. Because this peculiar structure is very common when dealing
with hospital admission, the algorithm of `augment()` takes this framework as a reference. So, what
does this mean? `pattern` can be either an integer, a factor or a character. Suppose we have it as 
an integer. `augment()` accepts only a pattern variable which have 2 or 3 unique values (i.e. 
running `length( unique( pattern ) )` must return 2 or 3). Now, suppose we provide a variable with
3 unique values. They must be 0, 1, and 3, nothing different than that. The coding for this is as
follows: 

1. Case 1. *integer*:
    + `pattern = 0`: subject is alive at the censoring time;
    + `pattern = 1`: subject is dead during a transition;
    + `pattern = 2`: subject is dead out of a transition.
1. Case 2. *factor*:
    + `pattern = 'alive'`: this is the first level of the factor and corresponds to `pattern = 0` 
    when integer;
    + `pattern = 'dead in'`: this is the second level of the factor and corresponds to `pattern = 1` 
    when integer;
    + `pattern = 'dead out'`: this is the third level of the factor and corresponds to `pattern = 2` 
    when integer;
1. Case 3. *character*:
    + the unique values must be in alphabetical order to resemble the pattern of the integer and 
    factor case.
    
Everything else to what described above will inevitably produce wrong behaviour of `augment()` with 
and uncorrect results.

The second important argument is `state`. This is passed as a list and contains the status flags 
which will be used to compute all the status variables for the process. The length of `state` is
3, no less, no more and comes with a default given by: `state = list( 'IN', 'OUT', 'DEAD' )`. 
The order is important here too. The status flags must be passed such that the first one (`'IN'`)
represents the first state (in `hosp`, being inside a hospital), the second one represents the 
second state (in `hosp`, being outside a hospital), and the third one represents the absorbing
state (in `hosp`, being dead inside or outside a hospital). As you can tell, this peculiar 
structure is typical of hospital admissions, where a patient can enter a hospital, can be discharged
from it, or can die. As we have already see, the `'DEAD'` status is reached no matter if the subject
has died inside or outside a transition (i.e. in our case, inside or outside the hospital).
One can change the flags in `state`, but they must be exactly 3.
One may need a higher level of complexity when specifying the states of subjects. This will be 
discussed in the next section where `state` acts as the main indicator of states.



## What if a more complex status structure is needed?

`augment()` by default takes a very simple status structure given by 3 different values. 
In general, this is enough to define a multi-state model. But what if we need a more complex
structure. Let’s consider again the dataset `hosp` for the 3rd, 4th, 5th, and 6th patient with the
following variables:

```{r subsetting, collapse = TRUE }
hosp[ 18:28, .( subj, adm_number, rehab, it, rehab_it,
                dateIN, dateOUT, dateCENS ) ]
```

As you can see, we have two variables which take into account the type of hospital admission.
*rehab* marks a rehabilitation admission while *it* marks an intensive therapy one.
They are both binary and integer variables, so one can compose them to get something which is
informative and, at the same time, usable in the context of ‘making a status’. We then created
the variable *rehab_it* which marks all the information in one place and it is a character.
You can pass *rehab_it* to the argument `more_status` to tell `augment()` to add these information into
a new structure. Now, it is important to remember that `augment()` introduces some rules when you
require to compute a more complex status structure. As you can see from the dataset, many values
of *rehab_it* are set to `df`. This stands for ‘default’ and when `augment()` finds it, it just compute
the default status you already passed to argument state (i.e. in this case, it can be ‘IN’, ‘OUT’,
or ‘DEAD’). The argument `more_status` always looks for the value `df`, hence whenever you need to
specify a default transition make sure to label it with this value. So, if we run `augment()` on
this sample, we obtain the following:

```{r complex status, collapse = TRUE }
hosp_augmented = augment( data = hosp, data_key = subj,
                          n_events = adm_number, pattern = label_2,
                          t_start = dateIN, t_end = dateOUT,
                          t_cens = dateCENS, more_status = rehab_it,
                          verbose = FALSE )

hosp_augmented[ 36:60, .( subj, adm_number, rehab_it,
                          augmented, status, status_exp, n_status_exp ) ]
```

Beside the usual status variables, of which we reported only status, `augment()` computed two more:

* *status_exp*: is the direct expansion of status and the variable you passed to `more_status`,
which in this case is *rehab_it*. The function composes them by pasting a ’_’ in between.
This is the main reason why it is worth to build a character variable if you know you need to fuel
it in as an indicator of a more complex status structure;
* *n_status_exp*: similar to what has been done before, `augment()` mixes information coming from
the current expandend status and the number of admission to give you the time evolution of the
process.


# Graphical Assessment of a Multi-state Model

**msmtools** has been mainly developed to easily manage and work with longitudinal datasets
which need to be restructured in order to get msm to work properly.

However, **msmtools** comes with two more functions which try to address graphically and in a very
efficient way the problem of the Goodness-of-Fit (Gof) for a multi-state model. When dealing
with this type of models, GoF is always a tough quest. Furthermore, up to now, no formal
statistical tests are defined when a multi-state model is computed within an exact time framework
[@titman2009model] [@titman2008general].

## Comparing fitted and empirical survival with `survplot()`

One of the most common graphical method to assess whether a multi-state model is behaving
the way we expect, is to compare the empirical survival with the fitted one. `survplot()`
helps out doing this and few more things. The function is a wrapper of the already known
`plot.survfit.msm()` from the package **msm**.

Suppose we ran a multi-state model on dataset hosp with the following code:

```{r multistate model, collapse = TRUE }
hosp_augmented = augment( data = hosp, data_key = subj,
                          n_events = adm_number, pattern = label_2,
                          t_start = dateIN, t_end = dateOUT,
                          t_cens = dateCENS, verbose = FALSE )

# let's define the initial transition matrix for our model
Qmat = matrix( data = 0, nrow = 3, ncol = 3, byrow = TRUE )
Qmat[ 1, 1:3 ] = 1
Qmat[ 2, 1:3 ] = 1
colnames( Qmat ) = c( 'IN', 'OUT', 'DEAD' )
rownames( Qmat ) = c( 'IN', 'OUT', 'DEAD' )
Qmat

# attaching the msm package and running the model using
# gender and age as covariates
library( msm )
msm_model = msm( status_num ~ augmented_int,
                 subject = subj, data = hosp_augmented,
                 covariates = ~ gender + age,
                 exacttimes = TRUE, gen.inits = TRUE,
                 qmatrix = Qmat, method = 'BFGS',
                 control = list( fnscale = 6e+05, trace = 0,
                                 REPORT = 1, maxit = 10000 ) )
```

We now have a multi-state model for which we can carry out some graphical inspections.
So, we want a simple comparison between the fitted survival curve and the empirical one,
computed using the Kaplan-Meier estimator. The code is as follows:

```{r survplot 1, fig.align = 'center', fig.width = 5, fig.height = 4}
survplot( msm_model, km = TRUE, ci = 'none',
          verbose = FALSE, devnew = FALSE )
```

With no surprises, the plot is not so satistying due to the really small dataset we provided.

Now, `survplot()` takes several parameters, many of them come with a default value. For instance, the
figureabove has been computed for a transition (IN - DEAD). We can pass to argument `from` any
starting state we want. If `to` is missing, `survplot()` will check what is the higher value in the
corresponding `msm` object and grabs it. Of course, you are free to compute any survival you want,
given the transition is allowed in the initial transition matrix `Qmat`. Let's plot the survival
comparison for the transition (OUT - DEAD):

```{r survplot 2, fig.align = 'center', fig.width = 5, fig.height = 4}
survplot( msm_model, km = TRUE, from = 2, ci = 'none',
          verbose = FALSE, devnew = FALSE )
```

If we do not want to show the Kaplan-Meier, we can pass `km = FALSE`, which is the default.

### Setting a custom time sequence

By default `survplot()` computes the fitted survival on a given grid. The number of grid points is
given by `grid`. In some cases, one would like to pass a custom time sequence. This can be achieved
by passing the argument `times` a numeric vector. Now `grid` is ignored.

Consider our dataset and suppose we want to compute a fitted survival only fo specific points in
time. The following code addresses this request.

```{r custom time seq, fig.align = 'center', fig.width = 5, fig.height = 4}
time_seq = seq( 300, 800, by = 30 )
survplot( msm_model, times = time_seq, ci = 'none',
          verbose = FALSE, devnew = FALSE )
```

### Obtaining the dataset for the Kaplan-Meier

It is possible to tell `survplot()` to return the associated Kaplan-Meier dataset by setting
`return.km = TRUE`. This fastly computes the data through `data.table`.
Passing only `km = TRUE` won't return any data, even if they must be computed anyway to plot results.

```{r returnKM, collapse = TRUE, fig.align = 'center', fig.width = 5, fig.height = 4}
survplot( msm_model, ci = 'none', return.km = TRUE,
          verbose = FALSE, do.plot = FALSE )
```

No results are printed. If you want to store the information in the current environment, 
you must assign `survplot()` to an object as follows:

```{r returnKM2, collapse = TRUE}
# running survplot() and assigning it to an object
km_data = survplot( msm_model, ci = 'none', return.km = TRUE,
                    verbose = FALSE, do.plot = FALSE )

# let's see the dataset
head( km_data )
```

The structure of the data is consistent. `survplot()` always computes a dataset in wide format, as
requested by `survfit` with 3 columns:

* *subject*: the ordered subject ID as passed to `msm` function;
* *mintime*: the time at which the event occurred;
* *anystate*: tansition indicator to compute the Kaplan-Meier.

The only modification you might encounter really depends on argument `exacttimes`.
This is inherited from `msm` function whose aim was to tell the model
that transitions occurred at exact and known times, including deaths. This is the main reason why
this argument should always be set the same way you set it in `msm`. In our case, we do have a
multi-state model in which transitions are well known and exact as you can see from the `msm` call
above. `survplot()` puts `exacttimes = TRUE` by default so we don't have to worry about it.
So, when `exacttimes = TRUE`, `survplot()` adds a new time variable which provides the relative
time for each transition within a given subject. For instance, `km_data` has another column 
named *mintime_exact* which is cast the same way of *augmented*.

### Obtaining the dataset for the fitted survival

Similarly to what done for the Kaplan-Meier, it is possible to obtain the data used to compute
the fitted survival as well. This can be achieve by setting `return.p = TRUE`. If `times` is passed,
then the resulting dataset will have as many rows as the elements in `times`. If `times` is missing,
then `survplot()` uses `grid` to know how many time points are requested. Below there is the
snippet that addresses what described.

```{r returnP, collapse = TRUE}
survplot( msm_model, ci = 'none', grid = 10, return.p = TRUE,
          verbose = FALSE, do.plot = FALSE )
```

As before, saving the data in the current environment follows the same procedure as seen before:

```{r returnP2, collapse = TRUE, fig.align = 'center', fig.width = 5, fig.height = 4}
# running survplot() and assigning it to an object
fitted_data = survplot( msm_model, ci = 'none', grid = 10, return.p = TRUE,
                        verbose = FALSE, do.plot = FALSE )

# let's see the dataset
fitted_data
```

The structure of the data is consistent here too. `survplot()` always computes a dataset
in wide format with 2 columns:

* *time*: time at which to compute the fitted survival. It can be obtained either by `grid`
or by `times` so that the cardinality of the data depends on them;
* *probs*: the corresponding value of the fitted survival.

Of course, you can request `survplot()` to return both the datasets by passing all the parameters.
Below you can see the code and the output when no assignment is done and when you save the data
into a new object.

```{r return_all, collapse = TRUE, fig.align = 'center', fig.width = 5, fig.height = 4}
# just running survplot()
survplot( msm_model, ci = 'none', grid = 10,
                     return.km = TRUE, return.p = TRUE,
                     verbose = FALSE, do.plot = FALSE )

# running survplot() and assigning it to an object
all_data = survplot( msm_model, ci = 'none', grid = 10,
                     return.km = TRUE, return.p = TRUE,
                     verbose = FALSE, do.plot = FALSE )

# let's see the dataset
all_data
```

`all_data` is a list of two elements. If you want to split up the datasets, just use common syntax:

```{r splitting data, collapse = TRUE}
# do not extract data using just one [].
# This keeps the class, so it returns a list
km_data_wrong = all_data[ 1 ]
# extracting data using the list way so be careful to use double []
km_data_1 = all_data[[ 1 ]]
# extracting data using the '$' access operator
km_data_2 = all_data$km
identical( km_data_wrong, km_data_1 )
identical( km_data_1, km_data_2 )
km_data_1

fitted_data_1 = all_data[[ 2 ]]
fitted_data_2 = all_data$fitted
identical( fitted_data_1, fitted_data_2 )
fitted_data_1
```

### The plot layering 

Multi-state models, as their name suggests, are specifically designed to work when more than one 
status exists. Infact, they tipically work when more than 2 status are present. In our example, 
we have 3 states: `'IN`, `'OUT'`, `'DEAD`. Now, what if we want to plot on the same device more
than just one fitted survival curve? Suppose we want to plot the expected survival computed from
state 1 and 2. That means, from `'IN` and `'OUT'` to `'DEAD'`. The plot layering addresses exactly
this problem and `survplot()` supports this with the argument `add`. Let's see this in the following
example: 

```{r layering, collapse = TRUE, fig.align = 'center', fig.width = 5, fig.height = 4}
# plotting the first survival from state 'IN'
survplot( msm_model, from = 1, devnew = FALSE, verbose = FALSE )

# plotting the first survival from state 'OUT'
survplot( msm_model, from = 2, add = TRUE, col.fit = 'darkblue',
          devnew = FALSE, verbose = FALSE )
```

## Comparing expected and observed prevalences with `prevplot()`

A second graphical tool which helps us in the attempt to understand the goodness of the model is
given by comparing the expected and observed prevalences. `prevplot()` is a wrapper of the
`plot.prevalence.msm()` function inside the `msm` package but, again, it does more things.

Consider the multi-state model we have built above. We can compute the prevalences using
`prevalence.msm()` function. This produces a named list which will be used inside `prevplot()`.
For instance, running the following code produces a plot of prevalences for each state of the
model.

```{r prev, collapse = TRUE, fig.align = 'center', fig.width = 7, fig.height = 3}
# defining the times at which compute the prevalences
t_min = min( hosp_augmented$augmented_int )
t_max = max( hosp_augmented$augmented_int )
steps = 100L

# computing prevalences
prev = prevalence.msm( msm_model, covariates = 'mean', ci = 'normal',
                            times = seq( t_min, t_max, steps ) )

# and plotting them using prevplot()
prevplot( msm_model, prev, ci = TRUE, devnew = FALSE, verbose = FALSE )
```

It is mandatory for `prevplot()` to work with a `msm` object and a list compute by `prevalence.msm`
are passed.

It is also possibile to plot the following statistic:

$$
M = \frac{(O_{is} - E_{is})^2} {E_{is}}
$$
which gives an idea of the deviance from the Markov model. This is computed according to Titman and
Sharples (2008). The following code addresses this request.

```{r plot_M, collapse = TRUE, fig.align = 'center', fig.width = 7, fig.height = 3, warning = F}
prevplot( msm_model, prev, M = TRUE, ci = TRUE, 
          devnew = FALSE, verbose = FALSE )
```


## A note on `verbose` and `devnew`

We are not going to discuss every arguments inside every function. This paragraph serves as a 
guideline to better exploit `msmtools`. We have developed the package by paying lot of attention to
the consistency of the code throughout the functions. 

For example, you can find the argument `verbose` anywhere in the package. This is useful when you 
do not want to be informed about anything that is happening internally with the only exception for 
warnings and of course errors. If quiet is your mood, then you may want to set `verbose = FALSE`.

A second argument which is shared between `survplot()` and `prevplot()` is `devnew`. This is a 
purely graphical parameter which is set to `TRUE` by default. It allows you to decide whether 
to plot on a new graphical device called with `dev.new()` by default. If you set `devnew = FALSE`, 
then the plot is overwritten according to `dev.cur()`. There exist two caveat: the former is that
`devnew` is bypassed if the argument `add = TRUE` is passed to `survplot()`. This means that, when
you want to add a new layer to an existing plot (with `add`), it does not make any sense to let 
the function using a new device. Hence, `devnew` is simply ignored. The latter is that, even you 
can set `devnew = FALSE` in `prevplot()`, one should keep it as the default. If you set it to `FALSE`
and you want a plot of `M` by setting `M = TRUE`, then this plot overwrites the prevalence one.





